Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-02 paper/ms.tex
update ms

\author{coauthors, possibly including but not limited to
Hillenbrand, Howard, Isaacson, Palumbo}

% <250 words
\begin{abstract}
  %Recent analyses of FGK stars in open clusters have revealed how
  %stellar rotation rates and lithium content evolve over the first few
  %billion years.
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium
  is measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid; these rotation-based ages are
  roughly an order of magnitude more precise than isochrone ages in
  this sample of FGK stars.
  %
  The independent lithium ages agree with the rotation-based ages for
  $\approx$90\% of cases, suggesting that most of the rotation-based
  ages are not only precise, they are also accurate.
  %However, the timescales of lithium depletion are such that
  %rotation-based ages are more precise for most of our stellar sample.
  %
  Our resulting catalog includes \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  More broadly, we find that the star formation rate in the Kepler
  field has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the
  past 3\,Gyr, equally for both planet hosts and for the overall
  stellar sample.
  %
  This ``demographic cliff'' is connected to the star formation
  history of the Galaxy, and it helps clarify the relative balance of
  young and old planets that have so far been detected using transits.
  %
\end{abstract}

@@

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
billions of years old, similar to the Earth.  This fact leaves many
gaps in our knowledge of what exact dynamical and physical processes
produce these objects.    The reason is that
processes such as thermal cooling \citep{2007ApJ...659.1661F},
atmospheric loss \citep{2019AREPS..47...67O}, giant impacts
\citep{2014prpl.conf..595R}, and dynamical instabilities
\citep{2017MNRAS.470.1750I} are expected to be most efficient over
$\approx$10\,Myr to $\approx$1\,Gyr timescales; when we observe most
known exoplanets, these processes have completed, and not much is happening.

Finding and characterizing young ($<$1\,Gyr) exoplanets is one 
possible path toward building a timeline for exoplanet evolution.
Informative individual young systems include V1298~Tau, an adolescent
resonant chain of four transiting puffy planets that is a likely
precusor to the compact multiplanet systems \citep{David_2019}, and
HIP~67522, a Jupiter-sized planet with a Neptunian mass
(\citealt{Rizzuto_2020}; Thao et al.~submitted).
% And perhaps also TOI837, now with a uber massive core!
Population-level analyses have similarly suggested differences in the
size distribution of young exoplanets relative to their older
counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

@@

To begin, imagine that the age distribution of stars within a few
hundred parsecs of the Sun is uniform between 0 and 10\,Gyr
\citep[e.g.][]{Nordstrom_2004}.  This approximation clarifies the
rarity of young stars: under these assumptions, $\approx$1\% of nearby
stars have $t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.
Entire fields focused on studying forming protoplanets
\citep{2018A&A...617A..44K}, and on watching exoplanets evolve shortly
after disk dispersal \citep[e.g.][]{2022MNRAS.512.5067K}, are thereby
capped in their maximum achievable sample sizes by the tyranny of the
Galactic star formation rate.

Despite the observational challenges, young close-in planet discovery
has done well over the past decade.  This has been primarily enabled
by Kepler, K2, TESS, and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The original strategy from around 2010 to 2020 was to focus on stars
with known ages---obvious members of open clusters---and to search
them for transiting planets.  The resulting stellar ages are precise
at the $\approx$10\% level, and planets discovered in this way are
generally assumed to have the same age as the host cluster.  A more
recent development, facilitated by the second Gaia data release, is
the idea that finding transiting planets can often be easier than
finding and characterizing the host stellar association
\citep[e.g.][]{Tofflemire_2021}.

Despite recent advances, the main problem with the ``cluster-first''
approach is that only $\approx$1\% of stars within the nearest few
hundred parsecs are currently associated with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
Most nearby sub-gigayear stars are in the field, and are omitted by
cluster-only planet searches.  
One can appreciate the problem by querying NASA's Exoplanet Archive
\citep[NEA;][]{2013PASP..125..989A}
for transiting planets younger than 1\,Gyr.  Imposing reasonable
precision requirements\footnote{For instance $t$$<$1\,Gyr at 2$\sigma$, or else
ages precise to say S/N$=$3.} gives $\approx$50 ``young''
transiting planets at the time of writing.  Most of these planets are
in clusters, and have been discovered by Kepler, K2, or TESS.
However, combined, those surveys have discovered $\approx$5{,}000
planets.  Unless there is a major bias against the discovery of
sub-gigayear old planets (with precisely constrained ages), we would
expect an order of magnitude more young transiting planets to be
known.

This study aims to resolve two main questions.  First, there is the
question of how wrong it is to assume a uniform age distribution for
stars in the galactic thin disk, particularly young stars.  Second,
where are the missing young transiting planets?  Briefly, we will find
that uniform is incorrect by a factor of a few, and that the missing
young transiting planets mostly have imprecise age measurements which
can be improved.

Given that Kepler's primary mission ended in 2013, one might be
forgiven for assuming that the ages of the Kepler planets are already
known.  Of course, they have been studied: \citet{Berger_2020b_rpage}
and \citet{Petigura_2022} for instance calculated isochrone ages for
Kepler Objects of Interest (KOIs).  These ages are most precise for
stars whose luminosities and temperatures separate them from the
main-sequence.  \citet{David_2021} leveraged both isochronal ages from
an earlier analogous study \citep{Fulton_Petigura_2018_cks_vii}, and
used a relative gyrochrone age-ranking approach that sorted stars
based on whether they rotated faster or slower than a gyrochrone age
of $\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.  These studies
were conducted in the broader context of investigating trends in
exoplanet demographics, which also depend on stellar metallicity and
mass \citep{Petigura_2018}.  These studies offered limited sensitivity
to the ages of the youngest planets, due to the nature of isochronal
dating.

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 0.5-4\,Gyr, and can also provide useful age limits at earlier
times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
clock is quite old
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  
In the context of Kepler,
empirical models have been previously applied to derive rotation-based
ages for subsets of the Kepler data
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.
There have been major improvements since these pioneering studies in:
\begin{enumerate}[label={\it \roman*)},leftmargin=*]
  \item The extent of the Kepler Object of Interest catalog, as well
    as our understanding of which of these objects are false
    positives;
  %
  \item Our empirical understanding of how rotation periods of stars
    in open clusters evolve
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item Our ability to model not only the mean evolution but also the
    intrinsic dispersion in rotation periods as a function of time
    \citep{Bouma_2023}.  
    For Sun-like stars, the new model is broadly consistent with older
    calibrations \citep[e.g.][]{Mamajek_2008}.  However, for late G
    dwarfs and K dwarfs, the age-calibration differs by factors of a
    few;
  %
  \item Our ability to identify binary stars, which can potentially
    problematic for estimates of rotation-based ages.
\end{enumerate}
Today, rotation-based ages for FGK stars below $\approx$4\,Gyr are
empirically tied to the cluster age scale and rotation sequence, and
are limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.

@@

{\it How do we use the bit-flags}
An important decision for our subsequent analysis is how to combine
the above quality indicators in order to decide whether or not a star is suited for
gyrochronology.  By default, we assume that a star is suitable for
gyrochronology if none of bits zero through nine (inclusive) are
raised.

@@

\begin{enumerate}[leftmargin=*]
  \item Two-sided posteriors available for both $t_{\rm gyro}$ and
    $t_{\rm Li}$: if their values were consistent within 2$\sigma$, we
    judged them to be consistent; if they were consistent within
    2-3$\sigma$, we suggest them to be ``maybe'' consistent.
    Otherwise, they are in disagreement.
    %
  \item If $t_{\rm Li}$ is a lower limit, we compare the $t_{\rm Li}$
    lower limit with the $1\sigma$ upper limit from $t_{\rm gyro}$; if
    the two overlapped, we judge the age estimates to be consistent.
    %
  \item If $t_{\rm Li}$ provides a two-sided posterior, and no
    rotation is reported, we judge the age estimates to be
    inconsistent.  This is because this lithium constraint can only be
    provided for G and K stars $\lesssim$2\,Gyr old, and Kepler should
    have been sensitive to the rotation periods of such stars.
\end{enumerate}

@@

With respect to our original questions, we have 
three main conclusions.
\begin{enumerate}[leftmargin=*]
  %FIXME: 120 number
  \item From the rotation rates of Kepler planet-hosting stars, we
    have identified  \nplyounggyrotwosigma\ planets younger than
    1\,Gyr at 2$\sigma$, and \nplyounggyro\ planets with median ages
    below 1\,Gyr.
    This implies that there are now $\approx$120-170 known
    sub-gigayear planets with well-constrained ages.
  %
  \item The age distributions of both the Kepler
    target stars and the known Kepler planet-hosts shows a
    demographic cliff.  
    There are \ratioobtoybstars\ times as many ``old'' (2-3\,Gyr)
    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.
    The star formation rate today is \ratiosfr$\pm$\uncratiosfr\ times
    lower than it was three billion years ago.
  %
  \item The implication of the latter age distribution is that rather
    than expecting $\approx$500 sub-gigayear exoplanets, we should
    instead expect $\approx$250.
    We have therefore gone from a factor of ten discrepancy to a
    factor of at most two.
%    The issue is therefore mostly, although perhaps not entirely, resolved.
%    Some of the remaining missing young
%    transiting planets may be in the Kepler data.  Many more are
%    likely in K2 and TESS.
\end{enumerate}

>>>>

\author{coauthors, including but not limited to
Hillenbrand, Howard, Isaacson, Palumbo}

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium
  is measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid; these rotation-based ages are
  roughly an order of magnitude more precise than isochrone ages in
  this sample of FGK stars.
  %
  The independent lithium ages agree with the rotation-based ages for
  $\approx$90\% of cases, suggesting that most of the rotation-based
  ages are not only precise, they are also accurate.
  %
  Our resulting catalog includes \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  More broadly, we find that the star formation rate in the Kepler
  field has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the
  past 3\,Gyr, equally for both planet hosts and for the overall
  stellar sample.
  %
  This ``demographic cliff'' is connected to the star formation
  history of the Galaxy, and it helps clarify the relative balance of
  young and old planets that have so far been detected using transits.
  %
\end{abstract}

@@

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
billions of years old, similar to the Earth.  This fact leaves many
gaps in our knowledge of what exact dynamical and physical processes
produce these objects.    The reason is that processes such as thermal
cooling \citep{2007ApJ...659.1661F}, atmospheric loss
\citep{2019AREPS..47...67O}, giant impacts
\citep{2014prpl.conf..595R}, and dynamical instabilities
\citep{2017MNRAS.470.1750I} are expected to be most efficient over
$\approx$10\,Myr to $\approx$1\,Gyr timescales; when we observe most
known exoplanets, most of the action is in the past.

Finding and characterizing young ($<$1\,Gyr) exoplanets is one
potential path toward building a timeline for exoplanet evolution.
Informative individual young systems include V1298~Tau, an adolescent
resonant chain of four transiting puffy planets that is a likely
precusor to the compact multiplanet systems \citep{David_2019}, and
HIP~67522, a Jupiter-sized planet with a Neptunian mass
(\citealt{Rizzuto_2020}; Thao et al.~submitted).  Population-level
analyses have similarly suggested differences in the size distribution
of young exoplanets relative to their older counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

@@

To begin, imagine that the ages of nearby stars in the Galaxy are
uniformly distributed between 0 and 10\,Gyr
\citep[e.g.][]{Nordstrom_2004}.  This approximation clarifies the
rarity of young stars: under these assumptions, $\approx$1\% of nearby
stars have $t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.
Studies of currently forming protoplanets \citep{2018A&A...617A..44K},
or of exoplanets evolving just after disk dispersal
\citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped in their
maximum achievable sample sizes by the tyranny of the Galactic star
formation rate.

Despite the observational challenges, young close-in planet discovery
has done well over the past decade, primarily due to
by Kepler, K2, TESS, and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The strategy pursued by most groups during the 2010s was to focus on
stars with known ages---obvious members of open clusters---and to
search them for transiting planets.  The resulting stellar ages are
precise at the $\approx$10\% level, and planets discovered in this way
are generally assumed to have the same age as the host cluster.  A
more recent development, facilitated by the second Gaia data release,
is the idea that finding transiting planets can often be easier than
finding and characterizing the host stellar association
\citep[e.g.][]{Tofflemire_2021}.

The main problem with the ``cluster-first'' approach is that most
nearby sub-gigayear stars are in the field.  Only $\approx$1\% of
stars within the nearest $\approx$500\,pc are currently associated
with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
One can appreciate the problem by querying NASA's Exoplanet Archive
\citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
than 1\,Gyr.  Imposing reasonable precision requirements\footnote{For
instance $t$$<$1\,Gyr at 2$\sigma$, or else ages precise to say
S/N$=$3.} gives $\approx$50 ``young'' transiting planets at the time
of writing.  Most of these planets are in clusters, and have been
discovered by Kepler, K2, or TESS.  However, combined, those surveys
have discovered $\approx$5{,}000 planets.  Unless there is a major
bias against the discovery of sub-gigayear old planets with precisely
constrained ages, we would expect an order of magnitude more young
transiting planets to be known.

This study aims to resolve two main questions.  First, how wrong it is
to assume a uniform age distribution for stars in the galactic thin
disk, particularly young stars?  Second, where are the missing young
transiting planets?  Briefly, we will find that uniform is incorrect
by a factor of a few, and that the bias against the discovery of young
transiting plantes has, to date, primarily been in the measurement of
theirs ages, rather than in the sensitivity of the transit searches.

Given that Kepler's primary mission ended in 2013, one might be
forgiven for assuming that the ages of the Kepler planets are already
known.  Of course, they have been studied: \citet{Berger_2020b_rpage}
and \citet{Petigura_2022} for instance calculated isochrone ages for
Kepler Objects of Interest (KOIs).  These ages are most precise for
stars whose luminosities and temperatures separate them from the
main-sequence.  \citet{David_2021} leveraged both isochronal ages from
an earlier analogous study \citep{Fulton_Petigura_2018_cks_vii}, and
used a relative gyrochrone age-ranking approach that sorted stars
based on whether they rotated faster or slower than a gyrochrone age
of $\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.  These studies
were conducted in the broader context of investigating trends in
exoplanet demographics, which also depend on stellar metallicity and
mass \citep[e.g.][]{Petigura_2018}.  These studies offered limited
sensitivity to the ages of the youngest planets, due to the nature of
isochronal dating.

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 0.5-4\,Gyr, and can also provide useful age limits at earlier
times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
clock has a rich history
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  In the
context of Kepler, empirical models have been previously applied to
derive rotation-based ages for subsets of the Kepler data
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  There have been
major improvements since these pioneering studies in:
\begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The extent of the Kepler Object of Interest catalog, as well
    as our understanding of which of these objects are false
    positives;
  %
  \item Our empirical understanding of how rotation periods of stars
    in open clusters evolve
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item Our ability to model not only the mean evolution but also the
    intrinsic dispersion in rotation periods as a function of time
    \citep{Bouma_2023}.  For Sun-like stars, the new model is broadly
    consistent with older calibrations \citep[e.g.][]{Mamajek_2008}.
    However, for late G dwarfs and K dwarfs, the age-calibration
    differs by factors of a few;
  %
  \item Our ability to identify binary stars, which can be problematic
    for estimates of rotation-based ages.
\end{enumerate}
Today, rotation-based ages for FGK stars below $\approx$4\,Gyr are
empirically tied to the cluster age scale and rotation sequence, and
are limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.

@@

{\it How do we use the bit-flags?}
An important decision for our subsequent analysis is how to combine
the above quality indicators in order to decide whether or not a star is suited for
gyrochronology.  By default, we assume that a star is suitable for
gyrochronology if none of bits zero through nine (inclusive) are
raised.

@@

\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item Two-sided posteriors available for both $t_{\rm gyro}$ and
    $t_{\rm Li}$: if their values were consistent within 2$\sigma$, we
    judged them to be consistent; if they were consistent within
    2-3$\sigma$, we suggest them to be ``maybe'' consistent.
    Otherwise, they are in disagreement.
    %
  \item If $t_{\rm Li}$ is a lower limit, we compare the $t_{\rm Li}$
    lower limit with the $1\sigma$ upper limit from $t_{\rm gyro}$; if
    the two overlapped, we judge the age estimates to be consistent.
    %
  \item If $t_{\rm Li}$ provides a two-sided posterior, and no
    rotation is reported, we judge the age estimates to be
    inconsistent.  This is because this lithium constraint can only be
    provided for G and K stars $\lesssim$2\,Gyr old, and Kepler should
    have been sensitive to the rotation periods of such stars.
\end{enumerate}

@@

Our 
three main conclusions with respect to our original questions are as follows.
\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  %FIXME: 120 number
  \item From the rotation rates of Kepler planet-hosting stars, we
    have identified  \nplyounggyrotwosigma\ planets younger than
    1\,Gyr at 2$\sigma$, and \nplyounggyro\ planets with median ages
    below 1\,Gyr.
    This implies that there are now $\approx$120-170 known
    sub-gigayear planets with well-constrained ages.
  %
  \item The age distributions of both the Kepler
    target stars and the known Kepler planet-hosts shows a
    demographic cliff.  
    There are \ratioobtoybstars\ times as many ``old'' (2-3\,Gyr)
    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.
    The star formation rate today is \ratiosfr$\pm$\uncratiosfr\ times
    lower than it was three billion years ago.
  %
  \item The implication of the latter age distribution is that rather
    than expecting $\approx$500 sub-gigayear exoplanets, we should
    instead expect $\approx$250.
    We have therefore gone from a factor of ten discrepancy to a
    factor of at most two.
%    The issue is therefore mostly, although perhaps not entirely, resolved.
%    Some of the remaining missing young
%    transiting planets may be in the Kepler data.  Many more are
%    likely in K2 and TESS.
\end{enumerate}
